---
title: "W9 Differential Gene Expression Analysis w DESeq2"
output: html_notebook
---

 
```{r}
library(DESeq2)
```
```{r}
library(dplyr)
```

```{r}
 install.packages("reshape2")
```
The count matrices for this analysis were downloaded from the NYU HPC and saved in a local folder titled source_files in this same working directory. 
#Create a DESeqDataSet object
```{r}
sampleFiles <- paste(c('PDAC253','PDAC282','PDAC286','PDAC316','PDAC266','PDAC273','PDAC306','PDAC318'),'.htseq_count.txt',sep="")
sampleCondition <- c(rep('highSucrose',4),rep('lowSucrose',4))
sampleTable <- data.frame(sampleName = sampleFiles,
                          fileName = sampleFiles,
                          condition = sampleCondition)
```
```{r}
sampleTable #Visualizing Sample Table 
```
```{r}
dds <- DESeqDataSetFromHTSeqCount(
  sampleTable = sampleTable,
  directory='/Users/danielaquijano/Documents/Github/Next-Generation-Sequence-Analysis/NGS_Course_Projects/Differential Gene Expression Analysis (W9)/source_files',
  design = ~ condition)
```
```{r}
dds
```
### 1.1
dds is a DESeq dataset, 28595 rows by 8 cols.

Test the null hypothesis that there is no effect of sugar type  on gene expression for each gene in the genome annotation. In the DESeq2 framework, we are testing the null hypothesis of a fold-change of zero separately for each gene in the genome between high sucrose and low sucrose fruits.

```{r}
class(dds)
```
## Pre-filter low count genes, normalize counts and run DESeq

```{r}
#removing genes with 10 or fewer reads
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
```


```{r}
#run the DESeq wrapper function. 

dds <- DESeq(dds)
class(dds) # Determine the type of object
```
Obtain raw counts and normalized counts from dds DESeq2 object.
```{r}
rawcounts.matrix <- counts(dds,normalized=F)
normalizedcounts.matrix <- counts(dds,normalized=T)
class(rawcounts.matrix)
```
```{r}
head(normalizedcounts.matrix)
```
### Q1.1
```{r}
dds
```

###2.1
As seen above, 28595 -20029=8566 genes were removed after filtering and normalization. 20029 genes were retained. 

##Analyze DGE data
Cluster samples using hierarchical clustering and Principal Components Analysis (PCA). This can help provide an indication of the distances between sample gene expression profiles and help detect potential batch effects.
Recommended that the normalized count matrix should be transformed with “regularized log” (?rlog) or “variance stabilizing” (?varianceStabilizingTransformation) transformation before clustering.
```{r}
# “rlog” transform the normalized counts and perform hierarchical clustering of samples:
rld <- rlog(dds)
dists <- dist(t(assay(rld)))
plot(hclust(dists))
```


```{r}
#Distance Matrix
dists
```


Cluster samples using the plotPCA function (?plotPCA) and color the points by high or low sucrose content samples 

```{r}
#Make PCA plot
plotPCA(rld)
```
###3.2a 
Samples are mainly clustered by sucrose content. sample _66 is an external node from the low sucrose group. Otherwise, the samples seem to cluster according to sucrose content. Sample _53 is a high sucrose sample but it clusters with other low sugar samples but it is still closer in relationship to high sugar then the low sugar samples below it (_73,_06,_18). 
On the PCA analysis, the samples cluster according to group. There is one sample from low sucrose group that is located away from low sucrose cluster and it is likely the same sample that was not clustered in the same clade in the dendogram.

### 3.2b
Yes. The PCA separates all points on the PC1 axis.
Note: Although there is one outlier in the far right, when we draw a vertical line we can still cleanly separate high and low sucrose groups despite that one point being vertically far away.

### 3.2c
Potential sources from batch effects:
all of the above: sequencer, lane, day, reagents, technicians

### 3.2d
Without any further evidence/statistical analysis I cannot say that there are any batch effects because I do not have information on how the samples were run and without any further analysis, or dimensionality breakdown other than PCA is is hard to identify batch effects. 

```{r}
#Results from the DESeq2 analysis can be extracted from the DESeqDataSet object using the results function.
res <- results(dds, contrast = c('condition','lowSucrose','highSucrose') )
class(res) 
```
```{r}
#sort the data ascending on p-value. 
#most significant differentially expressed gene at the top.
resOrdered <- res[order(res$pvalue),] 
head(resOrdered,10) # View the 10 most significant genes

```
DESeq2 identifies some of the genes displaying these problematic features by conducting “independent filtering”–an automated filtering process that causes p-values and/or adjusted p-values to be set to NA.

### 3.3 
Three quality control steps that are automatically conducted by DESeq to drop from consideration genes with suspect, or problematic, p-values 
How many genes were impacted by the default independent filtering?
-If all samples in a row have counts 0, then p value and adjusted p value will be set to NA.
-If a row has an outlier count, then p vlaue and adjusted value is set to NA
-If a row is filtered for low normalized count only adjusted p value is set to NA
In independent filtering the mean of the normalized counts are used as the filter statistic.8566 genes were removed after filtering and normalization.


The genes of interest in this experiment are three invertase genes that are located in a Quantitative Trait Locus (QTL) region approximately 1 Mb in size on chromosome 14 in the date palm genome.Invertase catalyzes the conversion of sucrose to fructose and glucose.



```{r}
library(pheatmap)
```

```{r}
#Filter genes from normalized counts using the top 30 most expressed genes
norm_counts_top_30=normalizedcounts.matrix[rownames(head(resOrdered,30)), ]
```


```{r}
#Here I arrange the annotation columns to generate an annotated and clsutered heatmap using pheatmap
annotation_columns=sampleTable
annotation_columns$fileName <- NULL
```
```{r}
annotation_columns
```
```{r}
row.names(annotation_columns) <- colnames(norm_counts_top_30)
```
```{r}
annotation_columns
```
```{r}
annotation_columns$sampleName <- NULL
```

```{r}
pheatmap(norm_counts_top_30, color=colorRampPalette(c("white", "light pink", "purple"))(30), scale="row", cluster_cols = T, show_rownames = T,fontsize_row = 4, fontsize_col = 4,labels_row = rownames(dists),main='Differentially Expressed genes with hierarchical clustering',annotation_col =annotation_columns )
```
The genes of interest in this experiment are three invertase genes that are located in a Quantitative Trait Locus (QTL) region approximately 1 Mb in size on chromosome 14 in the date palm genome. Invertase catalyzes the conversion of sucrose to fructose and glucose. We would like to know which (if any) of these genes are differentially expressed between sugar types and might control the sugar composition phenotype.<br>
The gene identifiers of the candidates are:<br>
Pdac_HC_chr14G0022900 (cell wall invertase enzyme)<br>
Pdac_HC_chr14G0023100 (cell wall invertase enzyme)<br>
Pdac_HC_chr14G0028200 (alkaline/neutral invertase enzyme)<br>
```{r}
#Subset sroted deseq results to find expression of candidate genes
sucrose_genes=resOrdered[ row.names(resOrdered) %in% c('Pdac_HC_chr14G0022900','Pdac_HC_chr14G0023100','Pdac_HC_chr14G0028200'), ]
sucrose_genes
```

### Q3.4a.
Based on the condition, the condition is in the format contrast=("cond","B","A") where it outputs log2(B/A). In this case B=low sucrose and A=high sucrose. Genes Pdac_HC_chr14G0023100 and Pdac_HC_chr14G0022900 were higher in the low sucrose sample whereas Pdac_HC_chr14G0028200  was higher in the high sucrose sample. In decimal form, the corresponding values for these log2 fold changes are:

```{r}
#Make a list of the candidate gene names:
candidate_genes<-c('Pdac_HC_chr14G0022900','Pdac_HC_chr14G0023100','Pdac_HC_chr14G0028200')
```

```{r}
decimal_value=exp(sucrose_genes$log2FoldChange)
original_log_change=sucrose_genes$log2FoldChange
log_to_dec_df=cbind(decimal_value,original_log_change)
```
```{r}
#Log to decimal value log fold change converting table. 
log_to_dec_df
```

### Q3.4B. report the normalized counts for each candidate gene as a table

```{r}
#normalized counts for each candidate gene as a table
plot_ct1<-plotCounts(dds, gene='Pdac_HC_chr14G0028200', intgroup="condition", returnData=TRUE)
```
```{r}
plot_ct2<-plotCounts(dds, gene='Pdac_HC_chr14G0023100', intgroup="condition", returnData=TRUE)
```
```{r}
plot_ct3<-plotCounts(dds, gene='Pdac_HC_chr14G0022900', intgroup="condition", returnData=TRUE)
```
```{r}
plot_ct1['gene name']<-rep(c('Pdac_HC_chr14G0028200'),each=8)
```
```{r}
plot_ct1['sample name']<-rownames(plot_ct1)
rownames(plot_ct1)<-NULL #Remove row names
```


```{r}
plot_ct2['gene name']<-rep(c('Pdac_HC_chr14G0023100'),each=8)
```
```{r}
plot_ct2['sample name']<-rownames(plot_ct2)
rownames(plot_ct2)<-NULL #Remove row names
```

```{r}
plot_ct3['gene name']<-rep(c('Pdac_HC_chr14G0022900'),each=8)
```
```{r}
plot_ct3['sample name']<-rownames(plot_ct3)
rownames(plot_ct3)<-NULL #Remove row names

```
```{r}
#Label the first column of each indidivual dataframe with the candidate gene name
names(plot_ct3)[1]='Pdac_HC_chr14G0022900'
names(plot_ct2)[1]='Pdac_HC_chr14G0023100'
names(plot_ct1)[1]='Pdac_HC_chr14G0028200'
```
```{r}
plot_ct1 #Sample dataframe
```
```{r}
plot_ct2
```
```{r}
plot_ct3
```

```{r}
#Make new dataframe by choosing columns from dataframes above
counts_candidates<-as.data.frame(cbind(plot_ct1['sample name'],plot_ct1['Pdac_HC_chr14G0028200'],plot_ct2['Pdac_HC_chr14G0023100'],plot_ct3['Pdac_HC_chr14G0022900'], plot_ct1['condition']))
```
```{r}
counts_candidates #This dataframe reflects the counts for all samples and all candidate genes
```
```{r}
library(tidyr)
library(reshape2)
```

```{r}
long <- melt(counts_candidates, id.vars = c("sample name", "condition"))
#Make dataframe into long format for graphing
```
```{r}
long
```

```{r}
names(long)[1]<-'sample' #Change column name to avoid spaces
```

```{r}
#Make a grouped bar plot with ggplot2 to show count data for candidate genes
ggplot(long, aes(x=variable, y=value, fill=sample)) + 
  geom_bar(stat="identity", position=position_dodge())+theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  ylab("Normalized Counts")+ xlab("Gene Candidate")+ggtitle("Counts for Candidate Invertase Genes") 
```
```{r}
#Make a grouped bar plot with ggplot2 to show count data for candidate genes
ggplot(long, aes(x=variable, y=value, fill=condition)) + 
  geom_bar(stat="identity", position=position_dodge())+theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  ylab("Normalized Counts")+ xlab("Gene Candidate")+ggtitle("Counts for Candidate Invertase Genes") 
```


```{r}
library(ggplot2)
library(reshape2)
```


DESEq2 log2 fold-change estimates are frequently over-estimated particularly for low expression genes. To obtain better estimates of the log2 fold-change, DESeq2 recommends “shrinking” the raw fold-change estimates DESeq2 makes available the lfcShrink function to extract results from a DESeqDataSet object with shrunken log2 fold-change estimates.
```{r}
library(devtools)
install_github("stephens999/ashr")
```

```{r}
res.shrunk <- lfcShrink(dds, contrast = c('condition','lowSucrose','highSucrose'),type='ashr' )
```

```{r}
res.shrunkOrdered <- res.shrunk[order(res.shrunk$pvalue),]
```

```{r}
resOrdered
```
```{r}
res.shrunkOrdered
```
### Q3.5 Use the plotMA function to generate MA plots for the res and res.shrunk objects. 
It is appropriate to report the shrunken estimates because genes with low expression can have high variability that can be reduced by shrinkage. 
```{r}
plotMA(res, ylim=c(-4,4))
```
```{r}
plotMA(res.shrunk , ylim=c(-4,4))
```

#Multiple Testing
Internally, DESeq2 will perform the False Discovery Rate (FDR) correction by running the p.adjust R core function (?p.adjust) on the raw p-values (pvalue column) using the Benjamin-Hochberg (“BH”) method. The output is reported in the “padj” column and is the adjusted p-value (i.e., the FDR). 

### Q4.1 Subset the table with “shrunken” log2 fold-change estimates to include only the candidate genes
### Q4.1a. Report results from the lfcShrink output table for the candidate genes.
```{r}
subset_res_shrunk<-res.shrunk[candidate_genes,]
subset_res_shrunk
```

###Q4.1b Using a criterion that a statistically differentially expressed gene must also show at least a two-fold change in expression (on linear scale) to be biologically meaningful, which genes do you consider to be differentially expressed? Please specify the gene(s), the FDR threshold you applied.
With FDR 0.001, the six  genes below can be consider differentially expressed since the amgnitude of the log fold change is greater than 2 and p value is less than 0.001.
```{r}
logfold_filter<-as.data.frame(subset(res.shrunk, res.shrunk[,2] > 2 |  res.shrunk[,2] < -2))
head(logfold_filter)
```


### Q4.1c Which of the candidate genes do you think could be responsible for the sugar composition trait on the chromosome 14 sugar QTL?
Based on log fold change, Pdac_HC_chr14G0022900 seems to be the best candidate based on the filtered table above. As seen on the bar plot above, it has the biggest ratio expression between low and high sucrose content. This means low sucrose palms express this gene lot whereas high sucrose palms express it minimally. 
